---
title: "DATA 607 - Assignment 2"
author: "Shoshana Farber"
date: "February 5, 2023"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(DBI)
library(dbplyr)
library(tidyverse)
library(RPostgres)
```

## Connecting to PostgreSQL Database

```{r}
my_pass <- read_file("C:/Users/Shoshana/Documents/pass.txt")

con <- dbConnect(
  Postgres(), 
  host = "localhost", 
  port = 5432,
  user = "postgres",
  password = my_pass, 
  dbname = "cuny-sps"
)
```

## Loading the Databases

```{r load-tables}
movie_ratings <- dbGetQuery(con, "SELECT * FROM movie_ratings")
raters <- dbGetQuery(con, "SELECT * FROM raters")
movies <- dbGetQuery(con, "SELECT * FROM movies")

# preview each table
head(movie_ratings)
head(raters)
head(movies)

# want a table of just the raters, movies, and their ratings 
movie_ratings <- movie_ratings %>%
  left_join(movies, on = "movieID") %>%
  left_join(raters, on = "raterID") %>%
  transmute(name, movie_title, rating)

head(movie_ratings)
```

## Exploration

### What is the average rating for each movie?

```{r avg-movie-ratings}
avg_rating <- movie_ratings %>%
  group_by(movie_title) %>% 
  filter(rating != is.na(rating)) %>%
  summarize(avg_rating = round(mean(rating), 2))

avg_rating
```

### Which movie is highest rated?

```{r highest-rated-movie}
avg_rating %>%
  arrange(-avg_rating)
```

Ticket to Paradise is the highest rated movie.

### How many people watched the highest rated movie?

```{r}
movie_ratings %>%
  filter(movie_title == "Ticket to Paradise", 
         rating != is.na(rating))
```

So ticket to paradise is the highest rated, but only 2/8 people actually watched it.

### Which movie did most people watch?

```{r most-watched}
watched <- movie_ratings %>%
  filter(rating != is.na(rating)) %>%
  mutate(watched = 1) 

watched %>%
  group_by(movie_title) %>%
  summarize(num_watched = sum(watched)) %>%
  arrange(-num_watched)
```

```{r graph-watches}
watched %>% 
  ggplot(aes(y = movie_title)) + 
    geom_bar(fill = "plum") +
    labs(title = "Total Watches per Movie")
```

Matilda the Musical and Top Gun Maverick were watched by the most people (7/8).

### What is the average rating given by each person?

```{r avg-person-rating}
avg_rating_person <- movie_ratings %>% 
  group_by(name) %>%
  filter(rating != is.na(rating)) %>%
  summarize(avg_rating = round(mean(rating), 2))

avg_rating_person
```

## Explanation

I connected to PostgreSQL server to access the `movie_ratings`, `movies`, and `raters` schemas directly from the database. `raters` has a primary key of **raterID**, and `movies` has a primary key of **movieID**. These are both foreign keys in `movie_ratings`. 

To analyze the data here, I reassigned `movie_ratings` to be a table of just the names of the raters, the movie titles, and their respective ratings. If I had more data, I would analyze popularity based on age, as the `raters` table includes the ages of those who rated. Additionally, all these movies are recent releases and were released in the same year, but if I had many more movies from many more years I would want to group based on year to see most popular movies per year. 
