---
title: "DATA 605 Homework 1"
author: "Shoshana Farber"
date: "January 29, 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("gifski")
library(gifski)
```

### Example - Building H

```{r}
x=c(rep(0,500),seq(0,1,length.out=1000), rep(1,500))
y=c(seq(-1,1,length.out=500),rep(0,1000), seq(-1,1,length.out=500))
z=rbind(x,y)
plot(y~x, xlim=c(-3,3), ylim=c(-3,3))
```

### Building my initials using plots in R

I modified the code for building an H in order to build my initials. 

```{r plot-both}
x <- c(rep(-2,500), seq(-2,-1,length.out=1000), rep(-1,500), seq(-2,-1,length.out=1000), seq(-2,-1,length.out=1000), seq(0,1,length.out=1000), rep(0, 500), seq(0,1,length.out=1000))
y <- c(seq(0,1,length.out=500), rep(0,1000), seq(-1,0, length.out=500), rep(1,1000), rep(-1,1000), rep(0,1000), seq(-1,1,length.out=500), rep(1,1000))
sf <- rbind(x,y)
plot(y~x, xlim=c(-3,2), ylim=c(-3,3))
```

### Write R code that will left multiply (%>%) a square matrix (x) against each of the vectors of points (y). Initially, that square matrix will be the Identity matrix. 

```{r}
leftMultiply <- function (x, y){
  x %*% y
}
```

### Use a loop that changes the transformation matrix incrementally to demonstrate 1) shear, 2) scaling, 3)  rotation , and 4) projection in animated fashion.

#### Shear

A shearing transformation causes the object to slant. For matrix [a,b,c,d], changing b or c leads to a shear transformation. Here I changed c to shear in the x direction. 

```{r animation.hook="gifski"}
for (i in seq(0,2,length.out=10)) {
  shear <- leftMultiply(matrix(c(1,i,0,1),nrow=2,ncol=2), sf)
  plot(shear[2,]~shear[1,], xlim=c(-3,2), ylim=c(-5,3))
}
```

#### Scaling

Scaling makes the object bigger or smaller. Here I scaled by the same value along the x and y axes.  

```{r animation.hook="gifski"}
for (i in seq(.5,1.5,length.out=10)) {
  scale <- leftMultiply(matrix(c(i,0,0,i),nrow=2,ncol=2), sf)
  plot(scale[2,]~scale[1,], xlim=c(-3,2), ylim=c(-3,3))
}
```

#### Rotation

Rotation causes the object to rotate around a point. Here, my initials rotate around the z-axis. 

```{r animation.hook="gifski"}
for (i in seq(0,2*pi,length.out=10)) {
  rotate <- leftMultiply(matrix(c(cos(i),-sin(i),sin(i),cos(i)),nrow=2,ncol=2), sf)
  plot(rotate[2,]~rotate[1,], xlim=c(-3,2), ylim=c(-3,3))
}
```

#### Projection

Projection flattens the image.

A rotation around the x-axis will flatten the image.

```{r animation.hook="gifski"}
project_sf <- rbind(sf, rep(0, ncol(sf)))

for (i in seq(0,2*pi,length.out=5)) {
  project <- leftMultiply(matrix(c(1,0,0,0,cos(i),-sin(i),0,sin(i),cos(i)), nrow=3, ncol=3), project_sf)
  plot(project[2,]~project[1,], xlim=c(-3,2), ylim=c(-3,3))
}
```
