---
title: "Homework 3"
author: "Shoshana Farber"
date: "February 12, 2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Set 1

(1) What is the rank of the matrix A?


$$A = \begin{bmatrix}1 & 2 & 3 & 4\\
-1 & 0 & 1 & 3\\
0 & 1 & -2 & 1\\
5 & 4 & -2 & -3
\end{bmatrix}$$

The rank of the matrix is equal to the number of non-zero rows of the matrix in reduced form:

```{r reduce-A}
A <- matrix(c(1,2,3,4,-1,0,1,3,0,1,-2,1,5,4,-2,-3), byrow=T, nrow=4)

pracma::rref(A)
```

The reduced row echelon form of this matrix is the 4x4 identity matrix, so the rank = 4.

(2) Given a mxn matrix where m > n, what is the maximum rank? What is the minimum rank,  assuming that the matrix is non-zero?

The maximum rank is the smaller of the amount of rows or columns. Since in this case m > n, the maximum rank of this matrix is n. Assuming this matrix is non-zero, the minimum rank is 1. 

(3)  What is the rank of matrix B?

$$B = \begin{bmatrix}1 & 2 & 1\\
3 & 6 & 3\\
2 & 4 & 2
\end{bmatrix}$$

We can see that row 2 and row 3 of this matrix are multiples of row 1 ($R_2 = 3R_1, R_3 = 2R_1$). Therefore, the rank of this matrix is 1.

## Problem Set 2

(1) Compute the eigenvalues and eigenvectors of the matrix A. You’ll need to show your work. You’ll need to write out the characteristic polynomial and show your solution.

$$A = \begin{bmatrix}1 & 2 & 3\\
0 & 4 & 5\\
0 & 0 & 6
\end{bmatrix}$$

### Finding the Characteristic Polynomial

$p_{A}(\lambda) = det(A-\lambda I_n)$
$$(A-\lambda I_n) = \begin{bmatrix}1-\lambda & 2 & 3\\
0 & 4-\lambda & 5\\
0 & 0 & 6-\lambda
\end{bmatrix}$$ 

The determinant of a triangular matrix is the product of the main diagonal:

$det(A-\lambda I_n) = (1-\lambda)(4-\lambda)(6-\lambda) = p_{A}(\lambda)$

### Finding the Eignevalues

By setting the characteristic polynomial equal to zero, we get the following eigenvalues:

$\lambda = 1, 4, 6$

### Finding the Eigenvectors

For $\lambda = 1$:

$$A-1I_n = \begin{bmatrix}0 & 2 & 3\\
0 & 3 & 5\\
0 & 0 & 5
\end{bmatrix} \underrightarrow{RREF}
\begin{bmatrix}0 & 1 & 0\\
0 & 0 & 1\\
0 & 0 & 0
\end{bmatrix}$$

```{r eval=FALSE}
pracma::rref(matrix(c(0,2,3,0,3,5,0,0,5), byrow=T, nrow=3))
```


$x_1$ is a free variable. From here we get a set of equations:
$x_1 = x_1$,
$x_2 = 0$,
$x_3 = 0$.

So, the corresponding eigenvector for $\lambda = 1$ is $\begin{bmatrix}1 \\ 0 \\ 0\end{bmatrix}$. 

For $\lambda = 4$:

$$A-1I_n = \begin{bmatrix}-3 & 2 & 3\\
0 & 0 & 5\\
0 & 0 & 2
\end{bmatrix} \underrightarrow{RREF}
\begin{bmatrix}1 & -\frac{2}{3} & 0\\
0 & 0 & 1\\
0 & 0 & 0
\end{bmatrix}$$

```{r eval=FALSE}
pracma::rref(matrix(c(-3,2,3,0,0,5,0,0,2), byrow=T, nrow=3))
```

$x_2$ is a free variable. From here we get a set of equations:
$x_1 -\frac{2}{3}x_2 = 0$,
$x_2 = x_2$,
$x_3 = 0$.

So, the corresponding eigenvector for $\lambda = 4$ is $\begin{bmatrix}\frac{2}{3} \\ 1 \\ 0\end{bmatrix}$. 

For $\lambda = 6$:

$$A-1I_n = \begin{bmatrix}-5 & 2 & 3\\
0 & -2 & 5\\
0 & 0 & 0
\end{bmatrix} \underrightarrow{RREF}
\begin{bmatrix}1 & 0 & -\frac{8}{5}\\
0 & 1 & -\frac{5}{2}\\
0 & 0 & 0
\end{bmatrix}$$

```{r eval=FALSE}
pracma::rref(matrix(c(-5,2,3,0,-2,5,0,0,0), byrow=T, nrow=3))
```

$x_3$ is a free variable. From here we get a set of equations:
$x_1 -\frac{8}{5}x_3 = 0$,
$x_2  -\frac{5}{2}x_3 = 0$,
$x_3 = x_3$.

So, the corresponding eigenvector for $\lambda = 6$ is $\begin{bmatrix}\frac{8}{5} \\ \frac{5}{2} \\ 1\end{bmatrix}$. 