---
title: "DATA 607 - Assignment 5"
author: "Shoshana Farber"
date: "February 26, 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(tidyverse)
library(snakecase) 
```

## Loading the Data

I loaded in the data, making sure to include `na.strings` so that any empty character cell would be filled with NA.

```{r}
untidy_flights <- read.csv(url("https://raw.githubusercontent.com/ShanaFarber/cuny-sps/master/DATA_607/Assignment5/flights.csv"), na.strings=c(""))

untidy_flights
```

## Cleaning

I made a `tidy_flights` data frame from `untidy_flights` and renamed the first two columns to be more descriptive.

```{r make-tidy-flights}
tidy_flights <- untidy_flights

tidy_flights <- tidy_flights %>% 
  rename("airline" = X,
         "arrival_status" = X.1)
```

I removed the empty row from the data frame and used `fill()` to populate the empty airline cells with their respective airline. 

```{r}
# remove empty row
tidy_flights <- tidy_flights[!apply(is.na(tidy_flights), 1, all),]

# populate empty airlines
tidy_flights <- tidy_flights %>% 
  fill(airline, .direction = "down")

tidy_flights
```

I found the code for removing the empty row [here](https://stackoverflow.com/questions/6437164/removing-empty-rows-of-a-data-file-in-r). 

Documentation on `fill()` function found [here](https://www.rdocumentation.org/packages/tidyr/versions/1.3.0/topics/fill). 

I then changed from wide to long format:

```{r}
# make long format
tidy_flights <- tidy_flights %>%
  pivot_longer(cols = c("Los.Angeles":"Seattle"),
               names_to = "city",
               values_to = "num_flights",
               values_transform = list(num_flights = as.character))

# remove any comma from num_flights and change to integer
tidy_flights$num_flights <- str_remove(tidy_flights$num_flights, ",")
tidy_flights$num_flights <- as.integer(tidy_flights$num_flights)

# remove . from city names
tidy_flights$city <- str_replace(tidy_flights$city, "\\.", " ")

# change arrival_status values to snake case
tidy_flights$arrival_status <- to_snake_case(tidy_flights$arrival_status)

tidy_flights
```

I then split based on "on time" vs. "delayed":

```{r}
tidy_flights <- tidy_flights %>%
  pivot_wider(names_from = "arrival_status",
              values_from = "num_flights")
```

## Comparing Arrival Delays

```{r}
tidy_flights <- tidy_flights %>%
  arrange(desc(delayed))

knitr::kable(tidy_flights)
```

```{r plot-delayed-flights}
tidy_flights %>%
  ggplot(aes(x = city, y = delayed, fill = airline)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = delayed), position = position_dodge(.9))
```

AM West flights to Pheonix had the most arrival delays (415 flights) while Alaska flights to Pheonix had the least arrival delays (12 flights). 

What was the proportion of delayed flights for each airline?

```{r prop-per-airline}
tidy_flights <- tidy_flights %>%
  mutate("total" = on_time + delayed)

delayed_per_airline <- tidy_flights %>% 
  group_by(airline) %>%
  summarize("delayed" = sum(delayed), "total" = sum(total)) %>%
  mutate("prop_delayed" = delayed/total) 

knitr::kable(delayed_per_airline)
```

About 13% of Alaska airline flights were delayed while about 11% of AM West flights were delayed. 

What was the proportion of delayed flights for each city?

```{r prop-per-city}
delayed_per_city <- tidy_flights %>%
  group_by(city) %>%
  summarize("delayed" = sum(delayed), "total" = sum(total)) %>%
  mutate("prop_delayed" = delayed/total) %>%
  arrange(desc(prop_delayed))

knitr::kable(delayed_per_city)
```

Flights to San Francisco had the most delays overall (22%). 

What was the proportion of delayed flights for each city according to each airline?

```{r prop-per-city-airline}
delayed_per_city_airline <- tidy_flights %>%
  mutate("prop_delayed" = delayed / total)

knitr::kable(delayed_per_city_airline)

delayed_per_city_airline %>%
  ggplot(aes(x = city, y = prop_delayed, color = airline)) +
    geom_point() +
    scale_y_continuous(labels = scales::percent)
```

We can see that AM West has a greater percentage of delayed flights for each city. AM West flights to San Francisco had the largest percentage of delayed flights (29%), while Alaska flights to Phoenix had the least (5%).

In all cases, it would seem that if I wanted to decrease my likelihood of being delayed to any city, I would want to take an Alaska airline flight. 




