---
title: "DATA 607 - Assignment 5"
author: "Shoshana Farber"
date: "February 26, 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(tidyverse)
library(snakecase) 
```

## Loading the Data

I loaded in the data, making sure to include `na.strings` so that any empty character cell would be filled with NA.

```{r}
untidy_flights <- read.csv(url("https://raw.githubusercontent.com/ShanaFarber/cuny-sps/master/DATA_607/Assignment5/flights.csv"), na.strings=c(""))

untidy_flights
```

## Cleaning

I made a `tidy_flights` data frame from `untidy_flights` and renamed the first two columns to be more descriptive.

```{r make-tidy-flights}
tidy_flights <- untidy_flights

tidy_flights <- tidy_flights %>% 
  rename("airline" = X,
         "arrival_status" = X.1)
```

I removed the empty row from the data frame and used `fill()` to populate the empty airline cells with their respective airline. 

```{r}
# remove empty row
tidy_flights <- tidy_flights[!apply(is.na(tidy_flights), 1, all),]

# populate empty airlines
tidy_flights <- tidy_flights %>% 
  fill(airline, .direction = "down")

tidy_flights
```

I found the code for removing the empty row [here](https://stackoverflow.com/questions/6437164/removing-empty-rows-of-a-data-file-in-r). 

Documentation on `fill()` function found [here](https://www.rdocumentation.org/packages/tidyr/versions/1.3.0/topics/fill). 

I then changed from wide to long format:

```{r}
# make long format
tidy_flights <- tidy_flights %>%
  pivot_longer(cols = c("Los.Angeles":"Seattle"),
               names_to = "city",
               values_to = "num_flights",
               values_transform = list(num_flights = as.character))

# remove any comma from num_flights and change to integer
tidy_flights$num_flights <- str_remove(tidy_flights$num_flights, ",")
tidy_flights$num_flights <- as.integer(tidy_flights$num_flights)

# remove . from city names
tidy_flights$city <- str_replace(tidy_flights$city, "\\.", " ")

# change arrival_status values to snake case
tidy_flights$arrival_status <- to_snake_case(tidy_flights$arrival_status)

tidy_flights
```

I then split based on "on time" vs. "delayed":

```{r}
tidy_flights <- tidy_flights %>%
  pivot_wider(names_from = "arrival_status",
              values_from = "num_flights")
```

## Comparing Arrival Delays

```{r}
tidy_flights <- tidy_flights %>%
  arrange(desc(delayed))

knitr::kable(tidy_flights)
```

```{r plot-delayed-flights}
tidy_flights %>%
  ggplot(aes(x = city, y = delayed, fill = airline)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = delayed), position = position_dodge(.9))
```

AM West flights to Pheonix had the most arrival delays (415 flights) while Alaska flights to Pheonix had the least arrival delays (12 flights). 

What was the proportion of delayed flights for each airline?

```{r prop-per-airline}
tidy_flights <- tidy_flights %>%
  mutate("total" = on_time + delayed)

delayed_per_airline <- tidy_flights %>% 
  group_by(airline) %>%
  summarize("delayed" = sum(delayed), "total" = sum(total)) %>%
  mutate("prop_delayed" = delayed/total) 

knitr::kable(delayed_per_airline)
```

About 13% of Alaska airline flights were delayed while about 11% of AM West flights were delayed. 

What was the proportion of delayed flights for each city?

```{r prop-per-city}
delayed_per_city <- tidy_flights %>%
  group_by(city) %>%
  summarize("delayed" = sum(delayed), "total" = sum(total)) %>%
  mutate("prop_delayed" = delayed/total) %>%
  arrange(desc(prop_delayed))

knitr::kable(delayed_per_city)
```

Flights to San Francisco had the most delays overall (22%). 

What was the proportion of delayed flights for each city according to each airline?

```{r prop-per-city-airline}
delayed_per_city_airline <- tidy_flights %>%
  mutate("prop_delayed" = delayed / total) %>%
  arrange(city, airline)

knitr::kable(delayed_per_city_airline)

delayed_per_city_airline %>%
  ggplot(aes(x = city, y = prop_delayed, color = airline)) +
    geom_point() +
    scale_y_continuous(labels = scales::percent)
```

We can see that AM West has a greater percentage of delayed flights for each city. AM West flights to San Francisco had the largest percentage of delayed flights (29%), while Alaska flights to Phoenix had the least (5%).

When comparing this to the overall percentage of delayed flights for each airline, it appears that Alaska Airlines outperformed AM West in every city, yet under-performed overall. This is because AM West has a greater overall number of flights relative to Alaska Airlines, and the calculation for delay percentage did not include a weighting of the delay percentage for each city relative to the percentage that each city's flights make of the total flights for each airline. 

Let us now perform the weighted calculations:

```{r}
# data frame for each airline's total flights
airline_totals <- tidy_flights %>%
  group_by(airline) %>%
  summarize("airline_total_flights" = sum(total))

# join to table of flight information
weighted_flights <- tidy_flights %>%
  left_join(airline_totals, on = "airline")

# calculate the percentage of flights to each city out of total airline flights
# calculate percentage of delayed flights out of total flights to city
weighted_flights <- weighted_flights %>%
  mutate("prop_of_flights" = total / airline_total_flights,
         "city_prop_delayed" = delayed / total)

# multiply percentages to get true percentage delay per airline to each city
weighted_flights <- weighted_flights %>%
  mutate("adj_prop_delayed" = prop_of_flights * city_prop_delayed)

# check that percentages still valid
weighted_flights %>% 
  group_by(airline) %>%
  summarize("prop_delayed" = sum(adj_prop_delayed))
```

The percentage of overall delays remains the same as above. We can now visualize the adjusted delay percentage of each airline's flights to each city:

```{r adj-delay-prop}
weighted_flights %>%
  ggplot(aes(x = city, y = adj_prop_delayed, color = airline)) +
    geom_point() +
    scale_y_continuous(labels = scales::percent)
```

```{r}
la_diff <- weighted_flights %>% 
  filter(city == "Los Angeles") %>%
  summarize(la_diff = diff(adj_prop_delayed))

la_diff <- la_diff$la_diff[1]
```

We can now compare the true proportion of delayed flights for each city. AM West outperforms Alaska Airlines in flights to San Francisco and Seattle, while Alaska outperforms AM West in flights to Phoenix and San Diego. The percentage of delayed flights to Los Angeles is similar between airlines, with AM West having a tiny edge on Alaska Airlines (a difference of about `r round(la_diff*100, 3)`%)

We can compare the two plots below to see how the percentage changes when the proportion is weighted correctly:

```{r fig.show='hold', out.width='50%'}
delayed_per_city_airline %>%
  ggplot(aes(x = city, y = prop_delayed, color = airline)) +
    geom_point() +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Delay Percentage of Each Airline Per City")

weighted_flights %>%
  ggplot(aes(x = city, y = adj_prop_delayed, color = airline)) +
    geom_point() +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Adjusted Delay Percentage of Each Airline Per City")
```




