---
title: "DATA 607 - Assignment 9"
author: "Shoshana Farber"
date: "March 26, 2023"
output: 
  rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)

library(tidyverse)
library(jsonlite)
library(kableExtra)
library(dotenv)

load_dot_env(".env")
```

**The New York Times web site provides a rich set of APIs, as described here: https://developer.nytimes.com/apis. You’ll need to start by signing up for an API key.**

**Your task is to choose one of the New York Times APIs, construct an interface in R to read in the JSON data, and transform it into an R DataFrame.**

## Connecting and Requesting from the API {.tabset .tabset-fade .tabset-pills}

I am using the **Top Stories** API health and science sections. 

```{r get-api}
url <- paste0('https://api.nytimes.com/svc/topstories/v2/science.json?api-key=', Sys.getenv("TIMES_API_KEY"))

url2 <- paste0('https://api.nytimes.com/svc/topstories/v2/health.json?api-key=', Sys.getenv("TIMES_API_KEY"))

science_data <- fromJSON(url)$results %>%
  as.data.frame() %>%
  add_column("web_section" = "Science", .before = "section")

health_data <- fromJSON(url2)$results %>%
  as.data.frame() %>%
  add_column("web_section" = "Health", .before = "section")
```

### Science 

```{r science-data-table}
rmarkdown::paged_table(science_data)
```

### Health

```{r health-data-table}
rmarkdown::paged_table(health_data)
```

## Tidying the Data {.tabset .tabset-fade .tabset-pills}

### Nested Format

In this format, I have removed columns that I do not want in the final data frame and I have left the `des_facet` and `org_facet` columns nested. 

```{r nested-format}
science_and_health_nested <- rbind(science_data, health_data) %>%
  select(web_section, title, abstract, url, "author" = byline, published_date, updated_date, des_facet, org_facet)

rmarkdown::paged_table(science_and_health_nested)
```

In this way, one has all the necessary information and can un-nest the `des_facet` or `org_facet` columns as they see fir for analysis. 

### Long Format

For the long format, the `des_facet` and `org_facet` columns are un-nested. The result is a set of data tables that have multiple rows for each article based on the number of tags in the `des_facet` or `org_facet` columns.  

```{r long-format}
# unnest des_facet
science_and_health_des_long <- rbind(science_data, health_data) %>%
  select(web_section, title, abstract, url, "author" = byline, published_date, updated_date, des_facet, org_facet) %>%
  unnest(des_facet)

#unnest org_facet
science_and_health_org_long <- rbind(science_data, health_data) %>%
  select(web_section, title, abstract, url, "author" = byline, published_date, updated_date, des_facet, org_facet) %>%
  unnest(org_facet)


rmarkdown::paged_table(science_and_health_des_long)

rmarkdown::paged_table(science_and_health_org_long)
```

In this format, the tags in the `des_facet` and `org_facet` columns can be analyzed. Alternatively, one could keep the nested data frame and un-nest the column that they wish to analyze.

### Short Format

For the short format, the `des_facet` and `org_facet` columns are unlisted and turned into a string in the data table. 

```{r short-format}
science_and_health <- rbind(science_data, health_data) %>%
  select(web_section, title, abstract, url, "author" = byline, published_date, updated_date, des_facet, org_facet)

############ unlisting des_facets #############
des_facets <- list()

for (i in 1:nrow(science_and_health)) {
  
  for (j in 1:length(unlist(science_and_health$des_facet[i]))) {
  temp <- paste(unlist(science_and_health$des_facet[i]), collapse='; ')
  
  }
  
  des_facets <- append(des_facets, list(temp))
}

des_facets <- data.frame(des_facets)

names(des_facets) <- c(1:length(names(des_facets)))

des_facets <- des_facets %>%
  pivot_longer(cols = names(des_facets), names_to = "article", values_to = "des_facet")

science_and_health <- science_and_health %>%
  mutate(des_facet = des_facets$des_facet)

############ unlisting org_facets #############
org_facets <- list()

for (i in 1:nrow(science_and_health)) {
  
  for (j in 1:length(unlist(science_and_health$org_facet[i]))) {
  temp <- paste(unlist(science_and_health$org_facet[i]), collapse='; ')
  
  }
  
  org_facets <- append(org_facets, list(temp))
}

org_facets <- data.frame(org_facets)

names(org_facets) <- c(1:length(names(org_facets)))

org_facets <- org_facets %>%
  pivot_longer(cols = names(org_facets), names_to = "article", values_to = "org_facet")

science_and_health <- science_and_health %>%
  mutate(org_facet = org_facets$org_facet)

rmarkdown::paged_table(science_and_health)
```

## Some Analysis

#### What are the most common descriptive tags? {.tabset .tabset-fade .tabset-pills}

The descriptive tags are in the `des_facet` column. For this, I will use the already un-nested `science_and_health_des_long` table.

##### Table

```{r des-facet}
most_common_des <- science_and_health_des_long %>% 
  count(des_facet) %>%
  arrange(desc(n)) %>%
  filter(n >= 5)
  
most_common_des %>%
  knitr::kable(col.names = c("Descriptive Tag", "Number of Articles"))
```

##### Graph

```{r graph-des-facets}
most_common_des %>%
  ggplot(aes(x = n, y = reorder(des_facet, n))) +
    geom_bar(stat = "identity", fill = "steelblue4") +
    geom_text(aes(label = n), position = position_stack(vjust = 0.9), fontface = 'bold', color = 'white') +
    labs(title = "Most Common Descriptive Tags", x = "", y = "")
```

"your-feed-science" is the most commonly used tag. Most articles seem to be about Covid-19 and other diseases. 

#### Which organizations are mentioned most? {.tabset .tabset-fade .tabset-pills}

For this, I will use the already un-nested `science_and_health_org_long` table.

##### Table

```{r org-facet}
most_common_org <- science_and_health_org_long %>% 
  count(org_facet) %>%
  arrange(desc(n)) %>%
  filter(n > 1)
  
most_common_org %>%
  knitr::kable(col.names = c("Organization", "Number of Articles"))
```
##### Graph

```{r graph-org-facets}
most_common_org %>%
  ggplot(aes(x = n, y = reorder(org_facet, n))) +
    geom_bar(stat = "identity", fill = "darkred") +
    geom_text(aes(label = n), position = position_stack(vjust = 0.9), fontface = 'bold', color = 'white') +
    labs(title = "Most Mentioned Organizations", x = "", y = "")
```

The FDA and CDC are the most mentioned organizations.

## Conclusions

The goal of this assignment was to in JSON data from the New York Times API and transform it into an R data frame. The "Tidying the Data" section provides three separate ways of organizing this data in an R data frame. 

In order to access the NYT API, follow the steps listed [here](https://developer.nytimes.com/get-started) on their website. 