---
title: "Homework 2"
author: "Shoshana Farber"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Set 1   

(1) Show that $A^{T}A \neq AA^{T}$ in general.

Matrix multiplication is not commutative. The order in which we multiply matrices matters, and for this reason $A^{T}A \neq AA^{T}$ generally, since $A^{T} \neq A$.

$$A = \begin{bmatrix}a_{11} & a_{12} & \cdots & a_{1n}\\
a_{21} & a_{22} &  & a_{2n}\\
\vdots & & \ddots & \vdots\\
a_{m1} & \cdots & & a_{mn}
\end{bmatrix}
\enspace
A^{T} = \begin{bmatrix}a_{11} & a_{21} & \cdots & a_{m1}\\
a_{12} & a_{22} &  & a_{m2}\\
\vdots & & \ddots & \vdots\\
a_{1n} & \cdots & & a_{mn}
\end{bmatrix}$$

We can see from here the result of $A^{T}A$ would be an nxn matrix while the result of $AA^{T}$ would be an mxm matrix. If $n \neq m$ then these matrices can obviously not be equal since they have different dimensions. 

For example, suppose we have $A$ and $A^{T}$ as follows:

$$A = \begin{bmatrix}a & b & c\\
d & e & f
\end{bmatrix}
\enspace
A^{T} = \begin{bmatrix}a & d\\
b & e\\
c & f
\end{bmatrix}$$

Since $A$ is a 2x3 matrix, and $A^{T}$ is a 3x2 matrix, the result of $A^{T}A$ will be a 3x3 matrix as follows:

$$A^{T}A = \begin{bmatrix}a & b & c\\
d & e & f
\end{bmatrix}
\begin{bmatrix}a & d\\
b & e\\
c & f\\
\end{bmatrix} = 
\begin{bmatrix}a^{2}+d^{2} & ab+de & ac+df\\
ba+ed & b^{2}+e^{2} & bc+ef\\
ca+fd & cb+fe & c^{2}+f^{2}
\end{bmatrix}$$

The result of $AA^{T}$ will likewise be a 2x2 matrix as follows:

$$AA^{T} = \begin{bmatrix}a & d\\
b & e\\
c & f\\
\end{bmatrix} 
\begin{bmatrix}a & b & c\\
d & e & f
\end{bmatrix}= 
\begin{bmatrix}a^{2}+b^{2}+c^{2} & ab+be+cf\\
da+eb+fc & d^{2}+e^{2}+f^{2}\\
\end{bmatrix}$$

We see here that $A^{T}A \neq AA^{T}$:

$$\begin{bmatrix}a^{2}+d^{2} & ab+de & ac+df\\
ba+ed & b^{2}+e^{2} & bc+ef\\
ca+fd & cb+fe & c^{2}+f^{2}
\end{bmatrix}
\neq
\begin{bmatrix}a^{2}+b^{2}+c^{2} & ab+be+cf\\
da+eb+fc & d^{2}+e^{2}+f^{2}\\
\end{bmatrix}$$

So, generally, $A^{T}A \neq AA^{T}$ as $A^{T} \neq A$. 


(2) For a special type of square matrix we get $A^{T}A = AA^{T}$. Under what conditions could this be true? (Hint: The Identity matrix $I$ is an example of such a matrix).

$A^{T}A = AA^{T}$ for symmetric square matrices such that $A^{T} = A$. This means that the rows of $A$ are the same as the columns of $A$, and as such, the transpose of matrix $A$ would remain $A$ itself. This would include diagonal square matrices, such as the identity matrix, as the transpose of any such matrix would be the same as the original matrix.  

$$\begin{bmatrix} a & b & c\\
b & d & e\\
c & e & f
\end{bmatrix}^{T} 
= 
\begin{bmatrix} a & b & c\\
b & d & e\\
c & e & f
\end{bmatrix}$$

## Problem Set 2

Write an R function to factorize a square matrix A into LU or LDU, whichever you
prefer. 

Here I figured out how to perform LU factorization on matrix A:

```{r}
A <- matrix(c(1,2,3,1,1,1,2,0,1),nrow=3)

U = A
L = diag(nrow(A))
  
for (i in 1:(nrow(A)-1)) {
  for (j in (i+1):nrow(A)) {
    L[j,i] = U[j,i]/U[i,i]
    U[j,] = U[j,] - L[j,i]*U[i,]
  }
}

L
U
```

When I first wrote the code, I had an error that indicated a value was out of range. I had to play around with the ranges for the for loops so I utilized the print function to walk me through what the code was doing at each step, as seen below: 

```{r eval=FALSE}
for (i in 1:(nrow(A)-1)) {
    print(i)
    for (j in (i+1):nrow(A)) {
      print(j)
      L[j,i] = U[j,i]/U[i,i]
      U[j,] = U[j,] - L[j,i]*U[i,]
      print(L)
      print(U)
    }
  }
```

I realized I had first had i as 1:nrow(A) and since j was i+1, there was a point where i=3 and j=4 for a 3x3 matrix so I changed i to 1:(nrow(A)-1). 

### The Completed Function

```{r}
luFactorize <- function(A) {
  U <-  A
  L <-  diag(nrow(A))
    
  for (i in 1:(nrow(A)-1)) {
    for (j in (i+1):nrow(A)) {
      L[j,i] = U[j,i]/U[i,i]
      U[j,] = U[j,] - L[j,i]*U[i,]
    }
  }
  
  LU <- list("L" = L, "U" = U)
  
  return(LU)
}
```

Testing the function to make sure it works properly: 

```{r}
test3x3 <- luFactorize(matrix(c(1,2,3,1,1,1,2,0,1), nrow=3))
test3x3$L %*% test3x3$U == matrix(c(1,2,3,1,1,1,2,0,1), nrow=3)

test4x4 <- luFactorize(matrix(c(1,2,3,1,1,1,2,0,1,4,6,2,0,5,1,3), nrow=4))
test4x4$L %*% test4x4$U == matrix(c(1,2,3,1,1,1,2,0,1,4,6,2,0,5,1,3), nrow=4)
```

