---
title: "DATA 605 - Homework 7"
author: "Shoshana Farber"
date: "March 12, 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**1.** Let $X_{1}$, $X_{2}$, . . . , $X_{n}$ be n mutually independent random variables, each of which is uniformly distributed on the integers from 1 to k. Let Y denote the minimum of the $X_{i}$’s. Find the distribution of Y.

For $Y = 1$: $P(Y=1)$ = ${{k^{n} - (k-1)^{n}} \over k^{n}}$ where $k^n$ is the total number of possibilities and $(k-1)^{n}$ is the possibilities where none of the $X_{i}$ are 1. 

For $Y = 2$: For $P(Y=2)$ = ${{(k-2)^{n} - (k-1)^{n} -[{k^{n} - (k-1)^{n}]}} \over k^{n}} = {{(k-1)^{n} - (k-2)^{n}} \over k^{n}}$. 

Following this pattern till $Y = y$:

$$
P(Y = y) = {{(k-y+1)^{n} - (k-y)^{n}} \over k^{n}} \space for \space 1<y<k
$$ 


**2.** Your organization owns a copier (future lawyers, etc.) or MRI (future doctors). This machine has a manufacturer’s expected lifetime of 10 years. This means that we expect one failure every ten years.

a) What is the probability that the machine will fail after 8 years? Provide also the expected value and standard deviation. Model as a **geometric**. (Hint: the probability is equivalent to not failing during the first 8 years.)

_**Formula for geometric:**_ $P(X = x) = (1-p)^{x-1} p$

```{r}
x <- 8
p_fail <- 1/10

1 - pgeom(x-1, p_fail)
```

$E(X) = p^{-1}$

```{r}
1/p_fail # expected value
```

$\sigma^2 = {1-p \over p^2}$

```{r}
var <- (1-p_fail) / (p_fail^2)

sqrt(var) #sd
```

The probability the machine fails after 8 years is **0.43 (43%)**. The expected failure time is **10 years**. The standard deviation is **9.49**. 

b) What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as an **exponential**. 

_**Formula for exponential:**_ $P(X = x) = {1 \over \theta} e^{-x \over \theta}$.

$\theta = {1 \over a}$

```{r}
a <- 1/10
x <- 8

1 - pexp(x, a)
```

$E(X) = \mu = \sigma$ for the exponential distribution. 

```{r}
theta <- 1/a

theta
```

The expected failure time is 10 years and the standard deviation is 10 years. 

c) What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as a **binomial**. (Hint: 0 success in 8 years) 

_**Formula for binomial:**_ $P(X = x) = {n \choose x} p^{x} q^{n-x}$

```{r}
x <- 0
n <- 8
p_fail = 1/10

pbinom(x, n, p_fail)
```

$E(X) = \mu = np$
$\sigma^{2} = npq$

```{r}
n*p_fail

n*p_fail*(1-p_fail)
```

The probability the machine fails after 8 years is **0.43 (43%)**. The expected number of failures is **0.8**. The standard deviation is **0.72**.

d) What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as a **Poisson**. 

_**Formula for binomial:**_ $P(X = x) = {{\lambda^{x} e^-\lambda} \over x!}$

```{r}
a <- 1/10
t <- 8

lambda <- a*t

ppois(0, lambda)
```

$E(X) = \mu = lambda$
$\sigma = \sqrt{\lambda}$

```{r}
lambda

sqrt(lambda)
```

The probability the machine fails after 8 years is **0.45 (45%)**. The expected number of failures is **0.8** and the standard deviation is **0.894**. 

