---
title: "DATA 607 - EC 1"
author: "Shoshana Farber"
date: "February 14, 2023"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(zoo)
library(scales)
```

### Loading the Data

Source: [Yahoo Finance BTC-USD](https://finance.yahoo.com/quote/BTC-USD/history?p=BTC-USD). 

The data set tracks the market price and volume of Bitcoin in USD. The data used was from January to December of 2022. 

```{r}
bitcoins <- read_csv(url("https://raw.githubusercontent.com/ShanaFarber/cuny-sps/master/DATA%20607/Extra%20Credit%201%20-%20Windows/BTC-USD.csv"), show_col_types = FALSE)

# change the type of 'Date' column to be a date
bitcoins$Date <- as.Date(bitcoins$Date)

knitr::kable(head(bitcoins))
```

### Computing the Average to Date and Six Day Rolling Window for Opening and Closing

```{r open}
bitcoins_open <- bitcoins %>% 
  transmute(date = Date, open = Open) %>%
  mutate(open_ytd_avg = cumsum(open) / seq_along(open),
         open_six_day_window = rollmean(open, 6, fill = NA))

knitr::kable(head(bitcoins_open))
```

```{r close}
bitcoins_close <- bitcoins %>%
  transmute(date = Date, close = Close) %>%
  mutate(close_ytd_avg = cumsum(close) / seq_along(close),
         close_six_day_window = rollmean(close, 6, fill = NA))

knitr::kable(head(bitcoins_close))
```

Information on "zoo" package and `rollmean()` function found [here](https://www.rdocumentation.org/packages/zoo/versions/1.8-11/topics/rollmean). 

I combined the two tables:

```{r join-open-close}
bitcoins_open_close <- inner_join(bitcoins_open, bitcoins_close, on = "Date")

knitr::kable(head(bitcoins_open_close))
```

### Some Visualization

I plotted the `open_ytd_avg` and `close_ytd_avg`. I used pivot longer to be able to plot both values on the same graph, differentiated by color. 

```{r plot-ytd-avgs}
bitcoins_open_close %>%
  pivot_longer(cols = c("open_ytd_avg", "close_ytd_avg"),
               names_to = "calculation",
               values_to = "price") %>%
  ggplot(aes(x = date, y = price, color = calculation)) +
  geom_line() +
  scale_y_continuous(labels = dollar_format())
```

You can see the decrease in the price of Bitcoin over the 2022 year. 

I also plotted the `six_day_window` against the daily closing price:

```{r plot-close-ytd-avg}
bitcoins_open_close %>%
  pivot_longer(cols = c("close", "close_six_day_window"),
               names_to = "calculation",
               values_to = "price") %>%
  ggplot(aes(x = date, y = price, color = calculation)) +
  geom_line() +
  scale_y_continuous(labels = dollar_format())
```

