---
title: "DATA 607 - Assignment 3"
author: "Shoshana Farber"
date: "February 12, 2023"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

### Problem 1

Using the 173 majors listed in fivethirtyeight.com’s [College Majors dataset](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/), provide code that identifies the majors that contain either "DATA" or "STATISTICS"

```{r load-majors}
majors <- read_csv(url("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv"), show_col_types = FALSE)

data_or_stat <- majors |>
  filter(str_detect(Major, "DATA") | str_detect(Major, "STATISTICS"))

knitr::kable(data_or_stat)
```

### Problem 2

Write code that transforms the data below:  

    [1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"  
    [5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
    [9] "elderberry"   "lime"         "lychee"       "mulberry"  
    [13] "olive"        "salal berry"  
    
Into a format like this:  

    c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

#### Loading the Data

```{r}
str1 <- '[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"' 
str2 <- '[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"'
str3 <- '[9] "elderberry"   "lime"         "lychee"       "mulberry"'
str4 <- '[13] "olive"        "salal berry"'

full_string <- paste(str1, str2, str3, str4)

full_string
```

#### Creating the Function

Initially I created the following function which replaces anything between the words with a comma, removes whatever is before the first entry or after the last entry, and then splits and formats the string:

```{r}
formatString <- function(string) {
  temp <- gsub('(\\[\\d+\\])?(\\s)?\\"', "", gsub('\"(\\s+)?(\\[\\d+\\])?(\\s+)?"', ",", string))
  
  temp <- unlist(str_split(temp, ","))
  
  dput(temp)
}

formatString(full_string)
```

I also created an alternative function which instead extracts all the desired words from the string and then formats the string properly:

```{r}
formatString <- function (string) {
  temp <- unlist(str_extract_all(string, '[a-zA-Z]+\\s?[a-zA_Z]+'))
  
  dput(temp)
}

formatString(full_string)
```

Both functions output the same desired format. 

### Problem 3

Describe, in words, what these expressions will match:

- `(.)\1\1`

      This will match any three consecutive characters that are all the same (ex: 'ooo'). The (.) captures a character in a group and each \1 calls that group (i.e. matches to that same character).
    
- `"(.)(.)\\2\\1"`

      This will match any four consecutive characters in the string where the last two characters are the same as the first two characters in reverse (ex: 'eppe' in pepper). Each (.) groups one characters and then \2 and \1 call those groups in the reverse.
      
- `(..)\1`

      This will match any four consecutive characters where the first two are the same as the last two (ex: 'anan' in banana). (..) groups two consecutive characters and \1 matches the same two characters.
      
- `"(.).\\1.\\1"`
  
      This will match any five consecutive characters where character 1, 3, and 5 are the same (ex: 'abana' in cabana). (.) groups a character, . can be any character, \1 matches the first character, . can again be any character, and \1 again matches the first character. 

- `"(.)(.)(.).*\\3\\2\\1"`

      This will match any string that contains at least six characters where the last three characters are the same as the first three characters in reverse order (ex: 'abcanythingcba'). Each (.) captures a different character, .* matches any number of characters, and \3, \2, \1 match the first three characters in reverse order. 
      
```{r test-regex-3}
test_string_3 <- c("coool", "banana", "cabana", "pepper", "boooo", "abcanythingcba", "deffed")

str_view(test_string_3, '(.)\\1\\1')

str_view(test_string_3, '(.)(.)\\2\\1')

str_view(test_string_3, '(..)\\1')

str_view(test_string_3, '(.).\\1.\\1')

str_view(test_string_3, '(.)(.)(.).*\\3\\2\\1')
```

### Problem 4

Construct regular expressions to match words that:

- Start and end with the same character.

  `"^(.).*\\1$"`

- Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice).
  
  `"(..).*\\1"`

- Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s).

  `"(.).*\\1.*\\1"`
  
```{r test-regex-4}
test_string_4 <- c("stats", "church", "blurb", "eleven", "strawberry", "pepper")

str_view(test_string_4, '^(.).*\\1$')

str_view(test_string_4, '(..).*\\1')

str_view(test_string_4, '(.).*\\1.*\\1')
```