---
title: "DATA 607 Week 1 - Basic Loading and Transformation"
author: "Shoshana Farber"
date: "January 29, 2023"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)

#install.packages("sqldf")
library(tidyverse)
library(sqldf)
library(scales)
```

## The Article

This assignment is based on the article _[According To Super Bowl Ads, Americans Love America, Animals And Sex](https://projects.fivethirtyeight.com/super-bowl-ads/)_ by FiveThirtyEight.com. This article categorizes Super Bowl commercials based on combinations of defining characteristics, such as whether there is a celebrity appearance, how patriotic the commercial is, etc. The authors of the article watched over 200 Super Bowl commercials and marked "yes" or "no" for each of the characteristics they were looking for. They then compared commercials that fit the same criteria and the ways in which advertisers chose to combine these criteria, sometimes with surprising results and drastically different approaches.

## Loading the Data

```{r load-data}
commercials <- read.csv(url("https://raw.githubusercontent.com/fivethirtyeight/superbowl-ads/main/superbowl-ads.csv"))

commercials$funny <- as.logical(commercials$funny)
commercials$show_product_quickly <- as.logical(commercials$show_product_quickly)
commercials$patriotic <- as.logical(commercials$patriotic)
commercials$celebrity <- as.logical(commercials$celebrity)
commercials$danger <- as.logical(commercials$danger)
commercials$animals <- as.logical(commercials$animals)
commercials$use_sex <- as.logical(commercials$use_sex)

commercials <- commercials %>% mutate(brand = replace(brand, brand == "Hynudai", "Hyundai"))
```

Documentation for the data set can be found [here](https://github.com/fivethirtyeight/superbowl-ads).

In the data set, the presence or absence of a defining characteristic in a commercial is denoted by character values "True" or "False." I chose to convert these to boolean (logical) values. 

I also noticed that the brand "Hyundai" was spelled wrong in the initial data set, so I used `mutate()` and `replace()` to change all instances of "Hynudai" to the correct spelling of "Hyundai."

## Analysis

```{r years}
commercials %>% summarize(start_year = min(year), end_year = max(year)) 
```

The years of the commercials span from 2000-2020.

```{r which-companies}
commercials %>%
  count(brand)
```

There are 10 brands which were studied in this article: Bud Light, Budweiser, Coca-Cola, Doritos, E-Trade, Hyundai, Kia, NFL, Pepsi, and Toyota. 

### How many commercials fall within each category?

```{r}
commercials %>% 
  summary()
```

In the summary, you can see the number of TRUE and FALSE observations for each column. 

|Column|True|False|
|--|--|--|
|funny|168|76|
|show_product_quickly|166|78|
|patriotic|41|203|
|celebrity|68|176|
|danger|75|169|
|animals|89|155|
|use_sex|63|181|

## Analyzing Combinations

Here I transformed the initial data set by adding columns for the combinations chosen by the article. 

Documentation for using SQL within R can be found [here](https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/sql.html). 

### Which commercials combine being funny with being patriotic?

```{r}
funny_and_patriotic <- sqldf("SELECT superbowl_ads_dot_com_url 
      FROM commercials 
      WHERE funny == TRUE and patriotic == TRUE") %>%
  mutate(funny_and_patriotic = TRUE)

funny_and_patriotic$superbowl_ads_dot_com_url

commercials <- commercials %>%
  left_join(funny_and_patriotic, by = "superbowl_ads_dot_com_url") %>%
  mutate(funny_and_patriotic = replace(funny_and_patriotic, is.na(funny_and_patriotic), FALSE))
```

### Which commercials combine danger with not being funny?

```{r}
danger_and_not_funny <- sqldf("SELECT superbowl_ads_dot_com_url 
      FROM commercials 
      WHERE danger == TRUE and funny == FALSE") %>%
  mutate(danger_and_not_funny = TRUE)

danger_and_not_funny$superbowl_ads_dot_com_url

commercials <- commercials %>%
  left_join(danger_and_not_funny, by = "superbowl_ads_dot_com_url") %>%
  mutate(danger_and_not_funny = replace(danger_and_not_funny, is.na(danger_and_not_funny), FALSE))
```

### Which commercials combine celebrities and being patriotic?

```{r}
celebrity_and_patriotic <- sqldf("SELECT superbowl_ads_dot_com_url 
      FROM commercials 
      WHERE celebrity == TRUE and patriotic == TRUE") %>%
  mutate(celebrity_and_patriotic = TRUE)

celebrity_and_patriotic$superbowl_ads_dot_com_url

commercials <- commercials %>%
  left_join(celebrity_and_patriotic, by = "superbowl_ads_dot_com_url") %>%
  mutate(celebrity_and_patriotic = replace(celebrity_and_patriotic, is.na(celebrity_and_patriotic), FALSE))
```

### Which commercials combine being funny, sex, and animals?

```{r}
funny_sex_and_animals <- sqldf("SELECT superbowl_ads_dot_com_url 
      FROM commercials 
      WHERE funny == TRUE and use_sex == TRUE and animals == TRUE") %>%
  mutate(funny_sex_and_animals = TRUE)

funny_sex_and_animals$superbowl_ads_dot_com_url

commercials <- commercials %>%
  left_join(funny_sex_and_animals, by = "superbowl_ads_dot_com_url") %>%
  mutate(funny_sex_and_animals = replace(funny_sex_and_animals, is.na(funny_sex_and_animals), FALSE))
```

```{r}
summary(commercials)
```

Adding to the table from above:

|Column|True|False|
|--|--|--|
|funny_and_patriotic|12|232|
|danger_and_not_funny|9|235|
|celebrity_and_patriotic|12|232|
|funny_sex_and_animals|20|224|

## Further Exploration

### Have commercials started to use sex more over the years?

```{r}
commercials %>% 
  group_by(year) %>%
  count(use_sex) %>%
  mutate(use_sex = tolower(as.character(use_sex))) %>%
  pivot_wider(names_from = "use_sex", values_from = "n") %>%
  mutate(true = replace(true, is.na(true), 0),
         total = true + false,
         prop = true / total) %>%
  ggplot(aes(x = year, y = prop)) +
    geom_smooth() +
    geom_point() +
    scale_y_continuous(labels = percent) +
    labs(title = "Sex in Super Bowl Ads Over the Years", 
         x = "Year", 
         y = "Proportion of Ads that Used Sex")
```

It actually seems that ads have used less sex over the years. 

### How many commercials did each brand make from 2000-2020?

To generate a table recording the number of ads created by each brand, we would use the following:

```{r}
commercials %>%
  count(brand)
```

If we wanted to visualize these results, we could use a bar graph to compare the number of commercials that each brand made from 2000-2020.

```{r}
commercials %>%
  ggplot(aes(x = brand)) +
    geom_bar(aes(color = brand, fill = brand)) +
    labs(title = "Number of Ads by Brand", 
         x = "Brand", 
         y = "Number of Ads") +
    theme(legend.position = "none")
```

From here we can see that Bud Light and Budweiser, both beer companies, made the most ads, with Bud Light making the most. We can also see that Hyundai made the most ads out of all three car companies, and Pepsi made a few more ads than their competitor, Coca-Cola. 

### What is the proportion of ads that show the product quickly?

```{r}
sum(as.integer(commercials$show_product_quickly)) / nrow(commercials)
```

68% of advertisements shows their products within the first 10 seconds.

### Which brands tend to show their product quickly?

```{r}
commercials %>%
  group_by(brand) %>%
  count(show_product_quickly) %>%
  mutate(show_product_quickly = tolower(as.character(show_product_quickly))) %>%
  pivot_wider(names_from = show_product_quickly, values_from = n) %>%
  mutate(total = false + true, prop = true / total)
```

Since some companies created more commercials than other, I used proportion to see which companies are quick to show their product. Doritos (88%), Toyota (81.8%), and Bud Light (81.7%) are the companies that are quickest to reveal themselves, with over 80% of their ads showing the product within the first 10 seconds. 

### Which brands combined humor, sex, and animals?

```{r}
commercials %>%
  group_by(brand) %>%
  count(funny_sex_and_animals) %>%
  mutate(funny_sex_and_animals = tolower(as.character(funny_sex_and_animals)))  %>%
  pivot_wider(names_from = funny_sex_and_animals, values_from = n)
```

Every company, except for Coca-Cola and the NFL, has in some way combined humor, sex, and animals in their advertisements. 

## Conclusions and Recommendations

Using a combination of R and SQL (from sqldf package), I was able to wrangle the data into a data frame that included the combinations mentioned in the article. Each smaller subset of the data frame, created using SQL code, would direct someone to the proper URL to be able to watch commercials based on their preferred combinations. Any of the code above in the "Analyzing Combinations" section can be modified to include new or alternative columns to create newly categorized ads based on new combinations. 

### Possible Extension

I would be interested to know which commercials and which combinations of characteristics are most popular. To do this, I would go through each YouTube link and add columns into the data set for `total_views` and `number_of_likes`. This way I could compare to see which advertisements are seem most popular and I could see what combination of factors were used. Alternatively, I could see which combinations are least popular based on most dislikes. 

This method would be somewhat limited, as not all people who enjoy a YouTube video actually take the time to "like" it. Also, most, if not all, of these advertisements are probably posted on many different YouTube channels while this data set only provides one. Therefore, any of the links provided in the data set could be less visited/liked than the same advertisement posted on another channel. However, I think this would anyways be a good start to measuring the popularity of these commercials, and it would certainly give insight to someone who was trying to figure out which of these criteria or combinations of factors would be good to put in their next ad. 