---
title: "Lab - Inference for Categorical Data"
author: "Shoshana Farber"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE, tidy.opts=list(width.cutoff=80), tidy=TRUE)

library(tidyverse)
library(openintro)
library(infer)

set.seed(613)
```

In this lab, we will explore and visualize the data using the **tidyverse** suite of 
packages, and perform statistical inference using **infer**. The data can be found
in the companion package for OpenIntro resources, **openintro**.

### The Data

We will be analyzing a sample from the Youth Risk Behavior Surveillance System (YRBSS) survey, which uses data from high schoolers to help discover health patterns. The data set is called `yrbss`.

```{r load-data}
data('yrbss', package='openintro')
```

### Exercise 1

**What are the counts within each category for the amount of days these students have texted while driving within the past 30 days?**

```{r text-while-drive}
text_while_driving <- yrbss |>
  count(text_while_driving_30d) |>
  mutate(text_while_driving_30d = replace(text_while_driving_30d, is.na(text_while_driving_30d), "unknown")) |>
  arrange(factor(text_while_driving_30d, levels = c("0", "1-2", "3-5", "6-9", "10-19", "20-29", "30", "did not drive", "unknown")))

text_while_driving |>
  knitr::kable(col.names = c("Days Texted While Driving (Out of 30)", "Number Reported"), format.args = list(big.mark = ","))

text_while_driving |>
  ggplot(aes(x = text_while_driving_30d, y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Days Texted While Driving (Out of 30)", y = "Number Reported", title = "Students Reporting Texting While Driving for X Days Within the Past 30 Days") +
  scale_y_continuous(labels = scales::comma)
```

### Exercise 2

**What is the proportion of people who have texted while driving every day in the past 30 days and never wear helmets?**

Remember that you can use `filter` to limit the dataset to just non-helmet wearers. Here, we will name the data set `no_helmet`.

```{r no-helmet}
no_helmet <- yrbss %>%
  filter(helmet_12m == "never") 

texted_everyday <- no_helmet |>
  count(text_while_driving_30d) |>
  mutate(total_students = sum(n),
         prop = n / total_students) |>
  filter(text_while_driving_30d == 30)
  
(texted_everyday <- paste0(round(texted_everyday$prop * 100, 2), "%"))
```

The proportion of students who texted everyday while driving and never wore helmets is **6.64%**. 

It may be easier to calculate the proportion if you create a new variable that specifies whether the individual has texted every day while driving over the past 30 days or not. We will call this variable `text_ind`.

```{r indicator-texting}
no_helmet <- no_helmet |>
  mutate(text_ind =ifelse(text_while_driving_30d == "30", "yes", "no"))

no_helmet |> 
  count(text_ind)

nrow(yrbss)

paste0(round((463/(6040 + 463 + 474))*100, 2), "%")
```

## Inference on Proportions

When summarizing the YRBSS, the Centers for Disease Control and Prevention seeks insight into the population *parameters*. To do this, you can answer the question, "What proportion of people in your sample reported that they have texted while driving each day for the past 30 days?" with a statistic; while the question "What proportion of people on earth have texted while driving each day for the past 30 days?" is answered with an estimate of the parameter.

The inferential tools for estimating population proportion are analogous to those used for means in the last chapter: the confidence interval and the hypothesis test.

```{r nohelmet-text-ci}
no_helmet |> 
  filter(!is.na(text_ind)) |>
  specify(response = text_ind, success = "yes") |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "prop") |>
  get_ci(level = 0.95)
```

Note that since the goal is to construct an interval estimate for a proportion, it's necessary to both include the `success` argument within `specify`, which accounts for the proportion of non-helmet wearers that have consistently texted while driving the past 30 days, in this example, and that `stat` within `calculate` is here "prop", signaling that you are trying to do some sort of inference on a proportion.

### Exercise 3

**What is the margin of error for the estimate of the proportion of non-helmet wearers that have texted while driving each day for the past 30 days based on this survey?**

```{r margin-of-error}
total_known_students <- no_helmet |>
  filter(!is.na(text_ind)) |>
  nrow()

p <- 463 / total_known_students

margin_error <- 1.96 * sqrt((p*(1-p))/total_known_students)

paste0(round(margin_error * 100, 2), "%")
```

The margin of error is about **0.63%**. 

### Exercise 4

**Using the `infer` package, calculate confidence intervals for two other categorical variables (you'll need to decide which level to call "success", and report the associated margins of error). Interpet the interval in context of the data.**

**Variable 1:** `physically_active_7d`

```{r}
physically_active_days <- yrbss |>
  count(physically_active_7d) |>
  mutate(physically_active_7d = replace(physically_active_7d, is.na(physically_active_7d), "unknown"))

physically_active_days |>
  knitr::kable(col.names = c("Days Per Week Physically Active", "Number Reported"), format.args = list(big.mark = ","))
```

**_What is the proportion of upperclassmen (juniors and seniors) who are active seven days a week?_**

```{r}
upperclassmen <- yrbss |>
  filter(grade %in% c(11,12)) |>
  mutate(active_ind =ifelse(physically_active_7d == "7", "yes", "no"))

upperclassmen %>% 
  filter(!is.na(active_ind)) |>
  specify(response = active_ind, success = "yes") |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "prop") |>
  get_ci(level = 0.95)

upperclassmen |>
  filter(!is.na(active_ind)) |>
  nrow()

upperclassmen |> 
  filter(active_ind == "yes") |>
  nrow()

p <- 1676 / 6619

margin_error <- 1.96 * sqrt((p*(1-p))/6619)

paste0(round(margin_error * 100, 2), "%")
```

**95% CI:** [0.243-0.264]
**ME:** 1.05%

We are 95% confident that the proportion of upperclassmen who are active seven days a week is between **24.3% and 26.4%**. 

**Variable 2:** `school_night_hours_sleep`

```{r}
hours_of_sleep <- yrbss |>
  count(school_night_hours_sleep) |>
  mutate(school_night_hours_sleep = replace(school_night_hours_sleep, is.na(school_night_hours_sleep), "unknown")) |>
  arrange(factor(school_night_hours_sleep, levels = c("<5", "5", "6", "7", "8", "9", "10+", "unknown")))

hours_of_sleep |>
  knitr::kable(col.names = c("Hours Slept Per School Night", "Number Students"), format.args = list(big.mark = ","))
```

**_What is the proportion of students who get more than eight hours of sleep on a school night?_**

```{r}
all_students <- yrbss |>
  mutate(sleep_ind =ifelse(school_night_hours_sleep %in% c("9", "10+"), "yes", "no"))

all_students %>% 
  filter(!is.na(sleep_ind)) |>
  specify(response = sleep_ind, success = "yes") |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "prop") |>
  get_ci(level = 0.95)

all_students |>
  filter(!is.na(sleep_ind)) |>
  nrow()

all_students |> 
  filter(sleep_ind == "yes") |>
  nrow()

p <- 1079/13583

margin_error <- 1.96 * sqrt((p*(1-p))/13583)

paste0(round(margin_error * 100, 2), "%")
```

**95% CI:** [0.075-0.084]
**ME:** 0.45%

We are 95% confident that the proportion of students who sleep more than eight hours on a school night is between **7.5% and 8.4%**. 

## How Does the Proportion Affect the Margin of Error?

Imagine you've set out to survey 1000 people on two questions: are you at least 6-feet tall? and are you left-handed? Since both of these sample proportions were calculated from the same sample size, they should have the same margin of error, right? Wrong! While the margin of error does change with sample size, it is also affected by the proportion.

Think back to the formula for the standard error: $SE = \sqrt{p(1-p)/n}$. This is then used in the formula for the margin of error for a 95% confidence interval: 

$$
ME = 1.96\times SE = 1.96\times\sqrt{p(1-p)/n} \,.
$$

Since the population proportion $p$ is in this $ME$ formula, it should make sense that the margin of error is in some way dependent on the population proportion. We can visualize this relationship by creating a plot of $ME$ vs. $p$.

Since sample size is irrelevant to this discussion, let's just set it to some value ($n = 1000$) and use this value in the following calculations:

```{r n-for-me-plot}
n <- 1000
```

The first step is to make a variable `p` that is a sequence from 0 to 1 with each number incremented by 0.01. You can then create a variable of the margin of error (`me`) associated with each of these values of `p` using the familiar approximate formula ($ME = 2 \times SE$).

```{r p-me}
p <- seq(from = 0, to = 1, by = 0.01)
me <- 2 * sqrt(p * (1 - p)/n)
```

Lastly, you can plot the two variables against each other to reveal their relationship. To do so, we need to first put these variables in a data frame that you can call in the `ggplot` function.

```{r me-plot, fig.show='hold', out.width='70%', fig.align='center'}
dd <- data.frame(p = p, me = me)
ggplot(data = dd, aes(x = p, y = me)) + 
  geom_line() +
  labs(x = "Population Proportion", y = "Margin of Error")
```

### Exercise 5

**Describe the relationship between `p` and `me`. Include the margin of error vs. population proportion plot you constructed in your answer. For a given sample size, for which value of `p` is margin of error maximized?**

The margin of error increases with the increase in population proportion up till 50% of the population. Then it begins to decrease again. 

## Success-Failure Condition

We have emphasized that you must always check conditions before making inference. For inference on proportions, the sample proportion can be assumed to be nearly normal if it is based upon a random sample of independent observations and if both $np \geq 10$ and $n(1 - p) \geq 10$. This rule of thumb is easy enough to follow, but it makes you wonder: what's so special about the number 10?

The short answer is: nothing. You could argue that you would be fine with 9 or that you really should be using 11. What is the "best" value for such a rule of thumb is, at least to some degree, arbitrary. However, when $np$ and $n(1-p)$ reaches 10 the sampling distribution is sufficiently normal to use confidence intervals and hypothesis tests that are based on that approximation.

You can investigate the interplay between $n$ and $p$ and the shape of the sampling distribution by using simulations. Play around with the following app to investigate how the shape, center, and spread of the distribution of $\hat{p}$ changes as $n$ and $p$ changes.

```{r sf-app, echo=FALSE, eval=FALSE}
library(shiny)
shinyApp(
  ui = fluidPage(
      numericInput("n", label = "Sample size:", value = 300),
      
      sliderInput("p", label = "Population proportion:",
                  min = 0, max = 1, value = 0.1, step = 0.01),
      
      numericInput("x_min", label = "Min for x-axis:", value = 0, min = 0, max = 1),
      numericInput("x_max", label = "Max for x-axis:", value = 1, min = 0, max = 1),
    plotOutput('plotOutput')
  ),
  
  server = function(input, output) { 
    output$plotOutput = renderPlot({
      pp <- data.frame(p_hat = rep(0, 5000))
      for(i in 1:5000){
        samp <- sample(c(TRUE, FALSE), input$n, replace = TRUE, 
                       prob = c(input$p, 1 - input$p))
        pp$p_hat[i] <- sum(samp == TRUE) / input$n
      }
      bw <- diff(range(pp$p_hat)) / 30
      ggplot(data = pp, aes(x = p_hat)) +
        geom_histogram(binwidth = bw) +
        xlim(input$x_min, input$x_max) +
        ggtitle(paste0("Distribution of p_hats, drawn from p = ", input$p, ", n = ", input$n))
    })
  },
  
  options = list(height = 500)
)
```

### Exercise 6

Describe the sampling distribution of sample proportions at $n = 300$ and $p = 0.1$. Be sure to note the center, spread, and shape.

The distribution appears normal with a center at 0.1. The distribution is narrow and tall so it is not very spread. There is a possible right skew. 

### Exercise 7

Keep $n$ constant and change $p$. How does the shape, center, and spread of the sampling distribution vary as $p$ changes. You might want to adjust min and max for the $x$-axis for a better view of the distribution.

As p moves closer to 0.5, the sample distribution becomes more normal and the spread decreases. As p increases from 0.5, the sample distribution, once again, gets more spread and develops a slight left skew. The center of the distribution increases corresponding to the increase in p. 
    
### Exercise 8

Now also change $n$. How does $n$ appear to affect the distribution of $\hat{p}$?

The sample size affects the spread of the distribution. A smaller sample size has a larger spread while a larger sample size has a smaller spread. 

* * *

## More Practice

For some of the exercises below, you will conduct inference comparing two proportions. In such cases, you have a response variable that is categorical, and an explanatory variable that is also categorical, and you are comparing the proportions of success of the response variable across the levels of the explanatory variable. This means that when using `infer`, you need to include both variables within `specify`.

### Exercise 9

Is there convincing evidence that those who sleep 10+ hours per day are more likely to strength train every day of the week? As always, write out the hypotheses for any tests you conduct and outline the status of the conditions for inference. If you find a significant difference, also quantify this difference with a confidence interval.

$H_{0}:$ Those who sleep 10+ hours a night are not more likely to strength train every day (there is no difference).

$H_{0}:$ Those who sleep 10+ hours a night are more likely to strength train every day (there is a difference). 

```{r}
ten_plus_hours_sleep <- yrbss |>
  filter(school_night_hours_sleep == "10+") |>
  mutate(train_ind = ifelse(strength_training_7d == 7, "yes", "no"))

ten_plus_hours_sleep |>
  filter(!is.na(train_ind)) |>
  count(train_ind) |>
  mutate(total = sum(n),
         prop = n / total)

yrbss |>
  mutate(train_ind = ifelse(strength_training_7d == 7, "yes", "no")) |>
  filter(!is.na(train_ind)) |>
  count(train_ind) |>
  mutate(total = sum(n),
         prop = n / total)
```

The proportion of students who sleep 10+ hours and train every day is **26.9%**. The proportion of all students who train every day is **16.8%**. The proportion of students who sleep 10+ hours and train every week is greater than the proportion of all students who train every day, so there is a significant difference and we can reject $H_{0}$. 

```{r}
ten_plus_hours_sleep |>
  filter(!is.na(train_ind)) |>
  specify(response = train_ind, success = "yes") |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "prop") |>
  get_ci(level = 0.95)

ten_plus_hours_sleep |>
  filter(!is.na(train_ind)) |>
  nrow()

ten_plus_hours_sleep |> 
  filter(train_ind == "yes") |>
  nrow()

p <- 84/312

margin_error <- 1.96 * sqrt((p*(1-p))/13583)

paste0(round(margin_error * 100, 2), "%")
```

**95% CI:** [0.221-0.317]

**ME:** 0.75%

We are 95% confident that the true proportion of students who sleep 10+ hours a night and train every day is between **22.1% and 31.7%**. The proportion found does fall within this confidence interval. 

### Exercise 10

Let's say there has been no difference in likeliness to strength train every day of the week for those who sleep 10+ hours. What is the probability that you could detect a change (at a significance level of 0.05) simply by chance?
  
*Hint:* Review the definition of the Type 1 error.

Type I error is the likelihood of rejecting a true null hypothesis (i.e. finding a significant difference when there is none). The probability of a type I error here is 5%. 

### Exercise 11

Suppose you're hired by the local government to estimate the proportion of residents that attend a religious service on a weekly basis. According to the guidelines, the estimate must have a margin of error no greater than 1% with 95% confidence. You have no idea what to expect for $p$. How many people would you have to sample to ensure that you are within the guidelines?

*Hint:* Refer to your plot of the relationship between $p$ and margin of error. This question does not require using a dataset.

The margin of error is greatest at 50% proportion. 

$ME = 1.96 \times SE$

$SE = \sqrt{p(1-p)/n}$

$0.01 = 1.96 \times \sqrt{0.5(1-0.5)/n}$

$0.01^{2} = 1.96^{2} \times 0.5(1-0.5)/n$

$n = 1.96^{2} \times 0.5(1-0.5)/0.01^{2}$

```{r}
(1.96**2) * (0.5*(1-0.5)) / (0.01**2)
```

You must be sure to sample at least **9,604** people to ensure that you are within the guidelines. 

* * *
