---
title: "Lab 3 - Probability"
author: "Shoshana Farber"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE, tidy.opts=list(width.cutoff=80), tidy=TRUE)

library(tidyverse)
library(openintro)
```

# The Hot Hand

Basketball players who make several baskets in succession are described as having a *hot hand*. Fans and players have long believed in the hot hand phenomenon, which refutes the assumption that each shot is independent of the next. However, [a 1985 paper](http://www.sciencedirect.com/science/article/pii/0010028585900106) by Gilovich, Vallone, and Tversky collected evidence that contradicted this belief and showed that successive shots are independent events. This paper started a great controversy that continues to this day, as you can see by Googling *hot hand basketball*.

We do not expect to resolve this controversy today. However, in this lab we'll apply one approach to answering questions like this. The goals for this lab are to (1) think about the effects of independent and dependent events, (2) learn how to simulate shooting streaks in R, and (3) to compare a simulation to actual data in order to determine if the hot hand phenomenon appears to be real.

## Load Data

We will focus on the performance of one player: [Kobe Bryant](https://en.wikipedia.org/wiki/Kobe_Bryant) of the Los Angeles Lakers. His performance against the Orlando Magic in the [2009 NBA Finals](https://en.wikipedia.org/wiki/2009_NBA_Finals) earned him the title *Most Valuable Player* and many spectators commented on how he appeared to show a hot hand. The data file we'll use is called `kobe_basket`.

```{r glimpse-data}
glimpse(kobe_basket)
```

This data frame contains 133 observations and 6 variables, where every row records a shot taken by Kobe Bryant. The `shot` variable in this dataset indicates whether the shot was a hit (`H`) or a miss (`M`).

Just looking at the string of hits and misses, it can be difficult to gauge whether or not it seems like Kobe was shooting with a hot hand. One way we can approach this is by considering the belief that hot hand shooters tend to go on shooting streaks. For this lab, we define the length of a shooting streak to be the *number of consecutive baskets made until a miss occurs*.

For example, in Game 1 Kobe had the following sequence of hits and misses from his nine shot attempts in the first quarter:

\[ \textrm{H M | M | H H M | M | M | M} \]

You can verify this by viewing the first 9 rows of the data in the data viewer.

```{r}
kobe_basket$shot[1:9]
```

Within the nine shot attempts, there are six streaks, which are separated by a "|" above. Their lengths are one, zero, two, zero, zero, zero (in order of occurrence).

### Exercise 1

**What does a streak length of 1 mean, i.e. how many hits and misses are in a streak of 1? What about a streak length of 0?**

A streak of length 1 means there was one hit before a miss. A streak of length 0 means there were zero hits. 

Counting streak lengths manually for all 133 shots would get tedious, so we'll use the custom function `calc_streak` to calculate them, and store the results in a data frame called `kobe_streak` as the `length` variable.

```{r calc-streak-kobe}
kobe_streak <- calc_streak(kobe_basket$shot)
```

We can then take a look at the distribution of these streak lengths.

```{r plot-streak-kobe}
ggplot(data = kobe_streak, aes(x = length)) +
  geom_bar()
```

### Exercise 2

**Describe the distribution of Kobe's streak lengths from the 2009 NBA finals. What was his typical streak length? How long was his longest streak of baskets?**

The distribution is right skewed. Kobe's typical streak was 0 to 1 hits, as those are the largest bins. His longest streak was 4 hits. 

## Compared to What?

We've shown that Kobe had some long shooting streaks, but are they long enough to support the belief that he had a hot hand? What can we compare them to?

To answer these questions, let's return to the idea of *independence*. Two processes are independent if the outcome of one process doesn't affect the outcome of the second. If each shot that a player takes is an independent process, having made or missed your first shot will not affect the probability that you will make or miss your second shot.

A shooter with a hot hand will have shots that are *not* independent of one another. Specifically, if the shooter makes his first shot, the hot hand model says he will have a *higher* probability of making his second shot.

Let's suppose for a moment that the hot hand model is valid for Kobe. During his career, the percentage of time Kobe makes a basket (i.e. his shooting percentage) is about 45%, or in probability notation,

\[ P(\textrm{shot 1 = H}) = 0.45 \]

If he makes the first shot and has a hot hand (*not* independent shots), then the probability that he makes his second shot would go up to, let's say, 60%,

\[ P(\textrm{shot 2 = H} \, | \, \textrm{shot 1 = H}) = 0.60 \]

As a result of these increased probabilities, you'd expect Kobe to have longer streaks. Compare this to the skeptical perspective where Kobe does *not* have a hot hand, where each shot is independent of the next. If he hit his first shot, the probability that he makes the second is still 0.45.

\[ P(\textrm{shot 2 = H} \, | \, \textrm{shot 1 = H}) = 0.45 \]

In other words, making the first shot did nothing to effect the probability that he'd make his second shot. If Kobe's shots are independent, then he'd have the same probability of hitting every shot regardless of his past shots: 45%.

Now that we've phrased the situation in terms of independent shots, let's return to the question: how do we tell if Kobe's shooting streaks are long enough to indicate that he has a hot hand? We can compare his streak lengths to someone without a hot hand: an independent shooter. 

## Simulations in R

While we don't have any data from a shooter we know to have independent shots, that sort of data is very easy to simulate in R. In a simulation, you set the ground rules of a random process and then the computer uses random numbers to generate an outcome that adheres to those rules. As a simple example, you can simulate flipping a fair coin with the following:

```{r head-tail}
coin_outcomes <- c("heads", "tails")
sample(coin_outcomes, size = 1, replace = TRUE)
```

If you wanted to simulate flipping a fair coin 100 times, you could either run the function 100 times or, more simply, adjust the `size` argument, which governs how many samples to draw:

```{r sim-fair-coin}
sim_fair_coin <- sample(coin_outcomes, size = 100, replace = TRUE)

sim_fair_coin
table(sim_fair_coin)
```

Since there are only two elements in `coin_outcomes`, the probability that we "flip" a coin and it lands heads is 0.5. Say we're trying to simulate an unfair coin that we know only lands heads 20% of the time. We can adjust for this by adding an argument called `prob`, which provides a vector of two probability weights:

```{r sim-unfair-coin}
set.seed(22)

sim_unfair_coin <- sample(coin_outcomes, size = 100, replace = TRUE, 
                          prob = c(0.2, 0.8))
```

### Exercise 3

**In your simulation of flipping the unfair coin 100 times, how many flips came up heads?**

```{r}
table(sim_unfair_coin)
```

In the simulation of the unfair coin, 16 flips came up heads.

## Simulating the Independent Shooter

Simulating a basketball player who has independent shots uses the same mechanism that you used to simulate a coin flip. To simulate a single shot from an independent shooter with a shooting percentage of 50% you can type:

```{r sim-basket}
shot_outcomes <- c("H", "M")
sim_basket <- sample(shot_outcomes, size = 1, replace = TRUE)
```

To make a valid comparison between Kobe and your simulated independent shooter, you need to align both their shooting percentage and the number of attempted shots.

### Exercise 4

**What change needs to be made to the `sample` function so that it reflects a shooting percentage of 45%? Make this adjustment, then run a simulation to sample 133 shots. Assign the output of this simulation to a new object called `sim_basket`.**

```{r}
set.seed(22)

sim_basket <- sample(shot_outcomes, size = 133, replace = TRUE, prob = c(0.45, 0.55))
```

With the results of the simulation saved as `sim_basket`, you have the data necessary to compare Kobe to our independent shooter.

Both data sets represent the results of 133 shot attempts, each with the same shooting percentage of 45%. We know that our simulated data is from a shooter that has independent shots. That is, we know the simulated shooter does not have a hot hand.

# More Practice

## Comparing Kobe Bryant to the Independent Shooter

### Exercise 5

**Using `calc_streak`, compute the streak lengths of `sim_basket`, and save the results in a data frame called `sim_streak`.**

```{r}
sim_streak <- calc_streak(sim_basket)
```

### Exercise 6

**Describe the distribution of streak lengths. What is the typical streak length for this simulated independent shooter with a 45% shooting percentage? How long is the player's longest streak of baskets in 133 shots? Make sure to include a plot in your answer.**

```{r plot-sim-streak}
ggplot(data = sim_streak, aes(x = length)) +
  geom_bar()
```

Similar to Kobe's, this distribution is right skewed. Typical streak length for this simulated independent shooter is 0. This player's longest streak is 4 baskets. 

### Exercise 7

**If you were to run the simulation of the independent shooter a second time, how would you expect its streak distribution to compare to the distribution from the question above? Exactly the same? Somewhat similar? Totally different? Explain your reasoning.**

Since I set the seed for the simulation shooter, I would expect the results to be exactly the same. If I had not set the seed, I would expect the results to be similar but to vary in length, distribution of streaks, and longest streaks, as the `sample()` function would take another random sampling. In fact, in running the simulation a few times without setting the seed, I can see that the samplings are similar in that there are usually roughly between 75-85 streaks long with most streaks being 0 or 1. However, they differ in how many of each streak there are, how long the longest streak is, and, as mentioned, the length of streaks. 

### Exercise 8

**How does Kobe Bryant's distribution of streak lengths compare to the distribution of streak lengths for the simulated shooter? Using this comparison, do you have evidence that the hot hand model fits Kobe's shooting patterns? Explain.**

```{r plot-sim-kobe-streak, fig.show='hold', out.width='50%'}
ggplot(data = kobe_streak, aes(x = length)) +
  geom_bar() +
  labs(title = "Kobe's Streaks")

ggplot(data = sim_streak, aes(x = length)) +
  geom_bar() +
  labs(title = "Simulation's Streaks")
```

We can see that the simulation's streaks are very similar to Kobe's, however Kobe has a few more 1-shot, 3-shot, and 4-shot streaks. I do not find evidence that the hot hand model fits Kobe's shooting patterns, as his pattern is similar to that of the independent simulation model. Therefore, I believe Kobe's shots are independent of one another and he does not have a "hot hand."
